|

### 2022-06-23 docs/rmd/paper.Rmd
autogit

---
title: |
  <center>Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression: </center>
  <center>
    A comparison of $\epsilon$-Lexicase Selection and Tournament Selection
  </center>
  <center>--------------------------------------------------------</center>
  <center>Student: Roman Höhn</center>
  <center>Date of Birth: 1991-04-14</center>
  <center>Place of Birth: Wiesbaden, Hesse</center>
  <center>Student ID: 2712497</center>
  <center>Supervisor: David Wittenberg</center>
  <center>--------------------------------------------------------</center>
  <center>Research Project for Seminar Information Systems (03.996.3299)</center>
  <center>FB 03: Chair of Business Administration and Computer Science</center>
  <center>Johannes Gutenberg University Mainz</center>
  <center>Summerterm 2022</center>
  |
date: "Date of Submission: `r Sys.Date()`"
output:
  pdf_document:
    keep_tex: true
    extra_dependencies: ["float"]
    fig_caption: yes
    number_sections: true
    includes:
      in_header: settings.tex
# indent: true
fontsize: 12pt
# toc: yes
# toc_depth: 4
bibliography: bib/bib/ref.bib
csl: bib/csl/harvard1.csl
# include-before: '`\newpage{} `{=latex}'
include-before: '`\thispagestyle{empty} \newpage{}`{=latex}'
---

@@

@@

Since regular Lexicase selection has been shown to perform sub optimal on continuous-valued optimization problems, a modified variation called $\epsilon$-Lexicase selection has been suggested for symbolic regression by La Cava et. al (2016). Here, $\epsilon$-Lexicase selection has shown itself to outperform both in overall performance while showing only negligible computational overhead [@epsilon_lexicase_main, p.747].

@@

<!-- Research Question --->
This research project tries to answer the question if the usage of $\epsilon$-Lexicase selection influences the generalization behavior of programs that are evolved using GP for symbolic regression if compared to programs that are evolved using traditional tournament selection.

@@

More recent research by @epsilon_lexicase_main suggests $\epsilon$-Lexicase selection as a modified selection operator that can improve overall GP performance if applied to continuous-valued symbolic regression tasks in comparison to tournament and standard lexicase selection.  @epsilon_lexicase_main argues that the original concept behind lexicase selection as formulated by @lexicase_first_desciption does not fit the requirements of real-world symbolic regression tasks. The authors identifies the pass condition of regular lexicase selection as the main problem if applied to symbolic regression: Individual solutions can only be selected if they perform on an elite level but with continuous-valued, and often noisy, data it is very unlikely for two individuals to achieve an exactly equal error for a given test case. This stringent pass condition can result in filtering out too many individuals during the selection process [@epsilon_lexicase_main, p.742]. $\epsilon$-Lexicase adresses this problem by introducing an $\epsilon$ parameter that specifies a range around the elite error, individuals that perform inside this range pass the current selection iteration. @https://doi.org/10.48550/arxiv.1709.05394 describes $\epsilon$-Lexicase selection as a more "relaxed version of lexicase selection".
Different methods to configure the $\epsilon$ Parameter have been explored, for the limited scope of this project I will focus on the most promising implementation of $\epsilon$-Lexicase selection that is automatically adjusted on the basis of the median absolute deviation of errors inside the selection pool [@epsilon_lexicase_main, p.742] [@https://doi.org/10.48550/arxiv.1709.05394, p.6].

The performance increase of $\epsilon$-Lexicase selection for symbolic regression problems has been demonstrated and reported for many benchmark problems which led to widespread adaption of it in symbolic regression applications [@epsilon_lexicase_main, p.744-745].

@@

The statistical analysis of the collected data will first focus on examining the question if, on average, the usage of $\epsilon$-Lexicase selection will result in models that perform significantly different than models that are evolved using tournament selection both.

@@

> $H0_{2}$: The distribution underlying the samples of training errors produced by $\epsilon$-Lexicase/torunament selection is the same as the distribution underlying samples of testing errors produced by $\epsilon$-Lexicase/tournament selection

@@

> $H0_{3}$: No differences in program size exist between the distribution underlying the samples produced by $\epsilon$-Lexicase and the distribution underlying samples of tournament selection

@@

Some interesting observation from tables 6 and 7 can be made by comparing the mean of the variables $min\_training\_error$ and $testing\_error$. Variable $min\_training\_error$ represents the fitness, measured by the MSE, of the best performing solution that has been found during a single run of the experiment while the variable $testing\_error$ represents the fitness of the same solution if evaluated for the unknown fitness cases inside the testing data set. Tournament selection based GP achieves a mean $min\_training\_error$ of $16.483$ and a mean $testing\_error$ of $16.905$ while the GP system using $\epsilon$-Lexicase selection can only achieve a mean $min\_training\_error$ of $22.013$ and a mean $testing\_error$ of $22.456$. These results indicate an overall performance benefit for tournament selection based GP in the experiment. It should also be mentioned that the best performing model over all 50 runs, measured by the minimum of variables $min\_training\_error$ and $testing\_error$, has been achieved by tournament selection with an MSE of 7.066 for the training data and	7.847 for the testing data.

If we compare the differences of $min\_training\_error$ and $testing\_error$ on the basis of each algorithm, we can observe that both tournament and $\epsilon$-lexicase achieve a slightly lower error on the training data than on the testing data. If we compute the total difference between the means of $min\_training\_error$ and $testing\_error$ for both algorithms tournament selection amounts to 0.422 and $\epsilon$-Lexicase selection to 0.443. Regarding our primary research question, the differences in generalization, the initial observations do not seem to indicate a significant difference in the the relative gap between the solutions performance on testing data and training data.

@@

1. The differences in mean fitness between tournament selection and $\epsilon$-Lexicase selection are highly statistical significant even if considering an $\alpha$ of 0.01. For the given symbolic regression task tournament selection achieves a better mean performance on both the training as well as the testing data set than $\epsilon$-Lexicase selection. These findings contradict the hypothesis $H0_{1}$ described in subsection 3.1 and are contrary to some of the research that was summarized in subsections 2.1 and 2.3.[@epsilon_lexicase_main] [@https://doi.org/10.48550/arxiv.1709.05394].

2. The differences in mean fitness between $min\_training\_error$ and $testing\_error$ are not statistically significant for both selection algorithms. If comparing the mean training error to the mean testing error, the p-values only computes to 0.309 for tournament selection and to 0.257 for $\epsilon$-Lexicase selection. These findings fail to reject $H0_{2}$, based on the obtained results we can not find evidence that both algorithms demonstrate a significantly different performance between the training data and the previously unseen testing data.

To further examine the generalization behavior of both selection operators I plotted the mean fitness scores for both algorithms over the total number of 100 generations in figure 2. Figure 2 shows that both algorithms show little visual evidence of over fitting which would result in a growing gap between between $min\_training\_error$ and $testing\_error$ in the final stages of the evolutionary optimization. For $\epsilon$-Lexicase selection a small gap can be observed starting at around the 90.th generation, while the minimum training error slowly converges a small increase in the testing error can be observed. Based on the limited amount of collected data, it is questionable if this small sign of over fitting can be interpreted as a weak indicator for an inferior generalization behavior of $\epsilon$-Lexicase in comparison to tournament selection.

Figure 2 also shows a visual representation of the described overall performance benefit of using tournament selection instead of $\epsilon$-Lexicase selection for the chosen symbolic regression task. It can be observed that at all stages of the evolution tournament selection based GP achieves a lower mean error for both data sets.

@@

None of the differences in mean size are statistically significant. Again, based on the results obtained in my experiment, the evidence produced fails to reject the null hypothesis $H0_{3}$ that seletion has no significant effect on differences in program size.

@@

...

@@

# I   References {.unnumbered #I}

@@

# II  Statutory Declaration {.unnumbered #II}

![Statutory Declaration](./misc/erklärung.pdf)\

>>>>

---
title: |
  <center>Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression: </center>
  <center>
    A comparison of $\epsilon$-lexicase Selection and Tournament Selection
  </center>
  <center>--------------------------------------------------------</center>
  <center>Student: Roman Höhn</center>
  <center>Date of Birth: 1991-04-14</center>
  <center>Place of Birth: Wiesbaden, Hesse</center>
  <center>Student ID: 2712497</center>
  <center>Supervisor: David Wittenberg</center>
  <center>--------------------------------------------------------</center>
  <center>Research Project for Seminar Information Systems (03.996.3299)</center>
  <center>FB 03: Chair of Business Administration and Computer Science</center>
  <center>Johannes Gutenberg University Mainz</center>
  <center>Summerterm 2022</center>
  |
date: "Date of Submission: `r Sys.Date()`"
output:
  pdf_document:
    keep_tex: true
    extra_dependencies: ["float"]
    fig_caption: yes
    number_sections: true
    includes:
      in_header: settings.tex
# indent: true
fontsize: 12pt
# toc: yes
# toc_depth: 4
bibliography: bib/bib/ref.bib
csl: bib/csl/harvard1.csl
# include-before: '`\newpage{} `{=latex}'
include-before: '`\thispagestyle{empty} \newpage{}`{=latex}'
---

@@

\listoftables
\listoffigures
\thispagestyle{empty}
\newpage

@@

Since regular Lexicase selection has been shown to perform sub optimal on continuous-valued optimization problems, a modified variation called $\epsilon$-lexicase selection has been suggested for symbolic regression by La Cava et. al (2016). Here, $\epsilon$-lexicase selection has shown itself to outperform both in overall performance while showing only negligible computational overhead [@epsilon_lexicase_main, p.747].

@@

<!-- Research Question --->
This research project tries to answer the question if the usage of $\epsilon$-lexicase selection influences the generalization behavior of programs that are evolved using GP for symbolic regression if compared to programs that are evolved using traditional tournament selection.

@@

More recent research by @epsilon_lexicase_main suggests $\epsilon$-lexicase selection as a modified selection operator that can improve overall GP performance if applied to continuous-valued symbolic regression tasks in comparison to tournament and standard lexicase selection.  @epsilon_lexicase_main argues that the original concept behind lexicase selection as formulated by @lexicase_first_desciption does not fit the requirements of real-world symbolic regression tasks. The authors identifies the pass condition of regular lexicase selection as the main problem if applied to symbolic regression: Individual solutions can only be selected if they perform on an elite level but with continuous-valued, and often noisy, data it is very unlikely for two individuals to achieve an exactly equal error for a given test case. This stringent pass condition can result in filtering out too many individuals during the selection process [@epsilon_lexicase_main, p.742]. $\epsilon$-lexicase adresses this problem by introducing an $\epsilon$ parameter that specifies a range around the elite error, individuals that perform inside this range pass the current selection iteration. @https://doi.org/10.48550/arxiv.1709.05394 describes $\epsilon$-lexicase selection as a more "relaxed version of lexicase selection".
Different methods to configure the $\epsilon$ Parameter have been explored, for the limited scope of this project I will focus on the most promising implementation of $\epsilon$-lexicase selection that is automatically adjusted on the basis of the median absolute deviation of errors inside the selection pool [@epsilon_lexicase_main, p.742] [@https://doi.org/10.48550/arxiv.1709.05394, p.6].

The performance increase of $\epsilon$-lexicase selection for symbolic regression problems has been demonstrated and reported for many benchmark problems which led to widespread adaption of it in symbolic regression applications [@epsilon_lexicase_main, p.744-745].

@@

The statistical analysis of the collected data will first focus on examining the question if, on average, the usage of $\epsilon$-lexicase selection will result in models that perform significantly different than models that are evolved using tournament selection both.

@@

> $H0_{2}$: The distribution underlying the samples of training errors produced by $\epsilon$-lexicase/torunament selection is the same as the distribution underlying samples of testing errors produced by $\epsilon$-lexicase/tournament selection

@@

> $H0_{3}$: No differences in program size exist between the distribution underlying the samples produced by $\epsilon$-lexicase and the distribution underlying samples of tournament selection

@@

Some interesting observation from tables 6 and 7 can be made by comparing the mean of the variables $min\_training\_error$ and $testing\_error$. Variable $min\_training\_error$ represents the fitness, measured by the MSE, of the best performing solution that has been found during a single run of the experiment while the variable $testing\_error$ represents the fitness of the same solution if evaluated for the unknown fitness cases inside the testing data set. Tournament selection based GP achieves a mean $min\_training\_error$ of $16.483$ and a mean $testing\_error$ of $16.905$ while the GP system using $\epsilon$-lexicase selection can only achieve a mean $min\_training\_error$ of $22.013$ and a mean $testing\_error$ of $22.456$. These results indicate an overall performance benefit for tournament selection based GP in the experiment. It should also be mentioned that the best performing model over all 50 runs, measured by the minimum of variables $min\_training\_error$ and $testing\_error$, has been achieved by tournament selection with an MSE of 7.066 for the training data and	7.847 for the testing data.

If we compare the differences of $min\_training\_error$ and $testing\_error$ on the basis of each algorithm, we can observe that both tournament and $\epsilon$-lexicase achieve a slightly lower error on the training data than on the testing data. If we compute the total difference between the means of $min\_training\_error$ and $testing\_error$ for both algorithms tournament selection amounts to 0.422 and $\epsilon$-lexicase selection to 0.443. Regarding our primary research question, the differences in generalization, the initial observations do not seem to indicate a significant difference in the the relative gap between the solutions performance on testing data and training data.

@@

1. The differences in mean fitness between tournament selection and $\epsilon$-lexicase selection are highly statistical significant even if considering an $\alpha$ of 0.01. For the given symbolic regression task tournament selection achieves a better mean performance on both the training as well as the testing data set than $\epsilon$-lexicase selection. These findings contradict the hypothesis $H0_{1}$ described in subsection 3.1 and are contrary to some of the research that was summarized in subsections 2.1 and 2.3.[@epsilon_lexicase_main] [@https://doi.org/10.48550/arxiv.1709.05394].

2. The differences in mean fitness between $min\_training\_error$ and $testing\_error$ are not statistically significant for both selection algorithms. If comparing the mean training error to the mean testing error, the p-values only computes to 0.309 for tournament selection and to 0.257 for $\epsilon$-lexicase selection. These findings fail to reject $H0_{2}$, based on the obtained results we can not find evidence that both algorithms demonstrate a significantly different performance between the training data and the previously unseen testing data.

To further examine the generalization behavior of both selection operators I plotted the mean fitness scores for both algorithms over the total number of 100 generations in figure 2. Figure 2 shows that both algorithms show little visual evidence of over fitting which would result in a growing gap between between $min\_training\_error$ and $testing\_error$ in the final stages of the evolutionary optimization. For $\epsilon$-lexicase selection a small gap can be observed starting at around the 90.th generation, while the minimum training error slowly converges a small increase in the testing error can be observed. Based on the limited amount of collected data, it is questionable if this small sign of over fitting can be interpreted as a weak indicator for an inferior generalization behavior of $\epsilon$-lexicase in comparison to tournament selection.

Figure 2 also shows a visual representation of the described overall performance benefit of using tournament selection instead of $\epsilon$-lexicase selection for the chosen symbolic regression task. It can be observed that at all stages of the evolution tournament selection based GP achieves a lower mean error for both data sets.

@@

None of the differences in mean size are statistically significant. Again, based on the results obtained in my experiment, the evidence produced fails to reject the null hypothesis $H0_{3}$ that selection has no significant effect on differences in program size. It should be noted that the configuration of both GP algorithm did include two mechanisms that place limitations on the growth behavior of the individual solutions to avoid excessive bloating (see sub section 3.2.4).

@@

The objective of this research project has been to answer the question if the choice between two commonly used parent selection operators, tournament selection and $\epsilon$-lexicase selection, significantly influences the generalization behavior of GP for symbolic regression tasks.

The results obtained from my experiment did not demonstrate any statistically significant differences in generalization, both GP systems achieve a slightly lower mean error if evaluated for the training data sets than if evaluated for the unseen fitness cases of the testing data set. The mean differences between testing and training errors did not show statistical significance differences for both algorithms (summarized in table 9).

To further explore differences in generalization behavior the mean for both algorithms training and testing error over all generations for all 50 independent runs was visualized and analyzed (see figure 2). A very small and probably insignificant gap did show for $\epsilon$-lexicase selection in the final phase of the evolutionary search, in the grand scheme of things this approach also did not yield any clear evidence of differences in generalization.

A final attempt to explore possible differences between both GP systems has been to explore differences in growth behavior as an indicator for generalization. This idea was based on the wide spread hypothesis that increases in overall program size might be an indication of over fitting to the training data, the concept has been discussed in sub section 2.2. Again, based on my results I could not find statistically significant differences in the growth behavior between both selection operators.

An interesting finding of my analysis is, that the overall performance of tournament selection has shown to be significantly better than that of $\epsilon$-lexicase selection for the selected symbolic regression problem. This finding was rather unexpected based on the research of @epsilon_lexicase_main and @https://doi.org/10.48550/arxiv.1709.05394 that primarily inspired this research project in the first place.

@@

# I - References {.unnumbered #I}

@@

# II - Statutory Declaration {.unnumbered #II}

![Statutory Declaration](./private/erklaerung.pdf)\

### 2022-06-23 docs/rmd/paper.tex
autogit

A comparison of \(\epsilon\)-Lexicase Selection and Tournament Selection

@@

Since regular Lexicase selection has been shown to perform sub optimal
on continuous-valued optimization problems, a modified variation called
\(\epsilon\)-Lexicase selection has been suggested for symbolic
regression by La Cava et. al (2016). Here, \(\epsilon\)-Lexicase
selection has shown itself to outperform both in overall performance
while showing only negligible computational overhead (La Cava, Spector
and Danai, 2016, p. 747).

@@

This research project tries to answer the question if the usage of
\(\epsilon\)-Lexicase selection influences the generalization behavior
of programs that are evolved using GP for symbolic regression if
compared to programs that are evolved using traditional tournament
selection.

@@

More recent research by La Cava, Spector and Danai (2016) suggests
\(\epsilon\)-Lexicase selection as a modified selection operator that
can improve overall GP performance if applied to continuous-valued
symbolic regression tasks in comparison to tournament and standard
lexicase selection. La Cava, Spector and Danai (2016) argues that the
original concept behind lexicase selection as formulated by Spector
(2012) does not fit the requirements of real-world symbolic regression
tasks. The authors identifies the pass condition of regular lexicase
selection as the main problem if applied to symbolic regression:
Individual solutions can only be selected if they perform on an elite
level but with continuous-valued, and often noisy, data it is very
unlikely for two individuals to achieve an exactly equal error for a
given test case. This stringent pass condition can result in filtering
out too many individuals during the selection process (La Cava, Spector
and Danai, 2016, p. 742). \(\epsilon\)-Lexicase adresses this problem by
introducing an \(\epsilon\) parameter that specifies a range around the
elite error, individuals that perform inside this range pass the current
selection iteration. La Cava \emph{et al.} (2017) describes
\(\epsilon\)-Lexicase selection as a more ``relaxed version of lexicase
selection''. Different methods to configure the \(\epsilon\) Parameter
have been explored, for the limited scope of this project I will focus
on the most promising implementation of \(\epsilon\)-Lexicase selection
that is automatically adjusted on the basis of the median absolute
deviation of errors inside the selection pool (La Cava, Spector and
Danai, 2016, p. 742) (La Cava \emph{et al.}, 2017, p. 6).

The performance increase of \(\epsilon\)-Lexicase selection for symbolic
regression problems has been demonstrated and reported for many
benchmark problems which led to widespread adaption of it in symbolic
regression applications (La Cava, Spector and Danai, 2016, pp.
744--745).

@@

The statistical analysis of the collected data will first focus on
examining the question if, on average, the usage of
\(\epsilon\)-Lexicase selection will result in models that perform
significantly different than models that are evolved using tournament
selection both.

@@

\begin{quote}
\(H0_{2}\): The distribution underlying the samples of training errors
produced by \(\epsilon\)-Lexicase/torunament selection is the same as
the distribution underlying samples of testing errors produced by
\(\epsilon\)-Lexicase/tournament selection
\end{quote}

@@

\begin{quote}
\(H0_{3}\): No differences in program size exist between the
distribution underlying the samples produced by \(\epsilon\)-Lexicase
and the distribution underlying samples of tournament selection
\end{quote}

@@

Some interesting observation from tables 6 and 7 can be made by
comparing the mean of the variables \(min\_training\_error\) and
\(testing\_error\). Variable \(min\_training\_error\) represents the
fitness, measured by the MSE, of the best performing solution that has
been found during a single run of the experiment while the variable
\(testing\_error\) represents the fitness of the same solution if
evaluated for the unknown fitness cases inside the testing data set.
Tournament selection based GP achieves a mean \(min\_training\_error\)
of \(16.483\) and a mean \(testing\_error\) of \(16.905\) while the GP
system using \(\epsilon\)-Lexicase selection can only achieve a mean
\(min\_training\_error\) of \(22.013\) and a mean \(testing\_error\) of
\(22.456\). These results indicate an overall performance benefit for
tournament selection based GP in the experiment. It should also be
mentioned that the best performing model over all 50 runs, measured by
the minimum of variables \(min\_training\_error\) and
\(testing\_error\), has been achieved by tournament selection with an
MSE of 7.066 for the training data and 7.847 for the testing data.

If we compare the differences of \(min\_training\_error\) and
\(testing\_error\) on the basis of each algorithm, we can observe that
both tournament and \(\epsilon\)-lexicase achieve a slightly lower error
on the training data than on the testing data. If we compute the total
difference between the means of \(min\_training\_error\) and
\(testing\_error\) for both algorithms tournament selection amounts to
0.422 and \(\epsilon\)-Lexicase selection to 0.443. Regarding our
primary research question, the differences in generalization, the
initial observations do not seem to indicate a significant difference in
the the relative gap between the solutions performance on testing data
and training data.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The differences in mean fitness between tournament selection and
  \(\epsilon\)-Lexicase selection are highly statistical significant
  even if considering an \(\alpha\) of 0.01. For the given symbolic
  regression task tournament selection achieves a better mean
  performance on both the training as well as the testing data set than
  \(\epsilon\)-Lexicase selection. These findings contradict the
  hypothesis \(H0_{1}\) described in subsection 3.1 and are contrary to
  some of the research that was summarized in subsections 2.1 and
  2.3.(La Cava, Spector and Danai, 2016) (La Cava \emph{et al.}, 2017).
\item
  The differences in mean fitness between \(min\_training\_error\) and
  \(testing\_error\) are not statistically significant for both
  selection algorithms. If comparing the mean training error to the mean
  testing error, the p-values only computes to 0.309 for tournament
  selection and to 0.257 for \(\epsilon\)-Lexicase selection. These
  findings fail to reject \(H0_{2}\), based on the obtained results we
  can not find evidence that both algorithms demonstrate a significantly
  different performance between the training data and the previously
  unseen testing data.
\end{enumerate}

To further examine the generalization behavior of both selection
operators I plotted the mean fitness scores for both algorithms over the
total number of 100 generations in figure 2. Figure 2 shows that both
algorithms show little visual evidence of over fitting which would
result in a growing gap between between \(min\_training\_error\) and
\(testing\_error\) in the final stages of the evolutionary optimization.
For \(\epsilon\)-Lexicase selection a small gap can be observed starting
at around the 90.th generation, while the minimum training error slowly
converges a small increase in the testing error can be observed. Based
on the limited amount of collected data, it is questionable if this
small sign of over fitting can be interpreted as a weak indicator for an
inferior generalization behavior of \(\epsilon\)-Lexicase in comparison
to tournament selection.

Figure 2 also shows a visual representation of the described overall
performance benefit of using tournament selection instead of
\(\epsilon\)-Lexicase selection for the chosen symbolic regression task.
It can be observed that at all stages of the evolution tournament
selection based GP achieves a lower mean error for both data sets.

@@

None of the differences in mean size are statistically significant.
Again, based on the results obtained in my experiment, the evidence
produced fails to reject the null hypothesis \(H0_{3}\) that seletion
has no significant effect on differences in program size.

@@

\ldots{}

@@

\hypertarget{I}{%
\section*{I References}\label{I}}
\addcontentsline{toc}{section}{I References}

@@

\hypertarget{II}{%
\section*{II Statutory Declaration}\label{II}}
\addcontentsline{toc}{section}{II Statutory Declaration}

\includegraphics{./misc/erklärung.pdf}\\

>>>>

A comparison of \(\epsilon\)-lexicase Selection and Tournament Selection

@@

Since regular Lexicase selection has been shown to perform sub optimal
on continuous-valued optimization problems, a modified variation called
\(\epsilon\)-lexicase selection has been suggested for symbolic
regression by La Cava et. al (2016). Here, \(\epsilon\)-lexicase
selection has shown itself to outperform both in overall performance
while showing only negligible computational overhead (La Cava, Spector
and Danai, 2016, p. 747).

@@

This research project tries to answer the question if the usage of
\(\epsilon\)-lexicase selection influences the generalization behavior
of programs that are evolved using GP for symbolic regression if
compared to programs that are evolved using traditional tournament
selection.

@@

More recent research by La Cava, Spector and Danai (2016) suggests
\(\epsilon\)-lexicase selection as a modified selection operator that
can improve overall GP performance if applied to continuous-valued
symbolic regression tasks in comparison to tournament and standard
lexicase selection. La Cava, Spector and Danai (2016) argues that the
original concept behind lexicase selection as formulated by Spector
(2012) does not fit the requirements of real-world symbolic regression
tasks. The authors identifies the pass condition of regular lexicase
selection as the main problem if applied to symbolic regression:
Individual solutions can only be selected if they perform on an elite
level but with continuous-valued, and often noisy, data it is very
unlikely for two individuals to achieve an exactly equal error for a
given test case. This stringent pass condition can result in filtering
out too many individuals during the selection process (La Cava, Spector
and Danai, 2016, p. 742). \(\epsilon\)-lexicase adresses this problem by
introducing an \(\epsilon\) parameter that specifies a range around the
elite error, individuals that perform inside this range pass the current
selection iteration. La Cava \emph{et al.} (2017) describes
\(\epsilon\)-lexicase selection as a more ``relaxed version of lexicase
selection''. Different methods to configure the \(\epsilon\) Parameter
have been explored, for the limited scope of this project I will focus
on the most promising implementation of \(\epsilon\)-lexicase selection
that is automatically adjusted on the basis of the median absolute
deviation of errors inside the selection pool (La Cava, Spector and
Danai, 2016, p. 742) (La Cava \emph{et al.}, 2017, p. 6).

The performance increase of \(\epsilon\)-lexicase selection for symbolic
regression problems has been demonstrated and reported for many
benchmark problems which led to widespread adaption of it in symbolic
regression applications (La Cava, Spector and Danai, 2016, pp.
744--745).

@@

The statistical analysis of the collected data will first focus on
examining the question if, on average, the usage of
\(\epsilon\)-lexicase selection will result in models that perform
significantly different than models that are evolved using tournament
selection both.

@@

\begin{quote}
\(H0_{2}\): The distribution underlying the samples of training errors
produced by \(\epsilon\)-lexicase/torunament selection is the same as
the distribution underlying samples of testing errors produced by
\(\epsilon\)-lexicase/tournament selection
\end{quote}

@@

\begin{quote}
\(H0_{3}\): No differences in program size exist between the
distribution underlying the samples produced by \(\epsilon\)-lexicase
and the distribution underlying samples of tournament selection
\end{quote}

@@

Some interesting observation from tables 6 and 7 can be made by
comparing the mean of the variables \(min\_training\_error\) and
\(testing\_error\). Variable \(min\_training\_error\) represents the
fitness, measured by the MSE, of the best performing solution that has
been found during a single run of the experiment while the variable
\(testing\_error\) represents the fitness of the same solution if
evaluated for the unknown fitness cases inside the testing data set.
Tournament selection based GP achieves a mean \(min\_training\_error\)
of \(16.483\) and a mean \(testing\_error\) of \(16.905\) while the GP
system using \(\epsilon\)-lexicase selection can only achieve a mean
\(min\_training\_error\) of \(22.013\) and a mean \(testing\_error\) of
\(22.456\). These results indicate an overall performance benefit for
tournament selection based GP in the experiment. It should also be
mentioned that the best performing model over all 50 runs, measured by
the minimum of variables \(min\_training\_error\) and
\(testing\_error\), has been achieved by tournament selection with an
MSE of 7.066 for the training data and 7.847 for the testing data.

If we compare the differences of \(min\_training\_error\) and
\(testing\_error\) on the basis of each algorithm, we can observe that
both tournament and \(\epsilon\)-lexicase achieve a slightly lower error
on the training data than on the testing data. If we compute the total
difference between the means of \(min\_training\_error\) and
\(testing\_error\) for both algorithms tournament selection amounts to
0.422 and \(\epsilon\)-lexicase selection to 0.443. Regarding our
primary research question, the differences in generalization, the
initial observations do not seem to indicate a significant difference in
the the relative gap between the solutions performance on testing data
and training data.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The differences in mean fitness between tournament selection and
  \(\epsilon\)-lexicase selection are highly statistical significant
  even if considering an \(\alpha\) of 0.01. For the given symbolic
  regression task tournament selection achieves a better mean
  performance on both the training as well as the testing data set than
  \(\epsilon\)-lexicase selection. These findings contradict the
  hypothesis \(H0_{1}\) described in subsection 3.1 and are contrary to
  some of the research that was summarized in subsections 2.1 and
  2.3.(La Cava, Spector and Danai, 2016) (La Cava \emph{et al.}, 2017).
\item
  The differences in mean fitness between \(min\_training\_error\) and
  \(testing\_error\) are not statistically significant for both
  selection algorithms. If comparing the mean training error to the mean
  testing error, the p-values only computes to 0.309 for tournament
  selection and to 0.257 for \(\epsilon\)-lexicase selection. These
  findings fail to reject \(H0_{2}\), based on the obtained results we
  can not find evidence that both algorithms demonstrate a significantly
  different performance between the training data and the previously
  unseen testing data.
\end{enumerate}

To further examine the generalization behavior of both selection
operators I plotted the mean fitness scores for both algorithms over the
total number of 100 generations in figure 2. Figure 2 shows that both
algorithms show little visual evidence of over fitting which would
result in a growing gap between between \(min\_training\_error\) and
\(testing\_error\) in the final stages of the evolutionary optimization.
For \(\epsilon\)-lexicase selection a small gap can be observed starting
at around the 90.th generation, while the minimum training error slowly
converges a small increase in the testing error can be observed. Based
on the limited amount of collected data, it is questionable if this
small sign of over fitting can be interpreted as a weak indicator for an
inferior generalization behavior of \(\epsilon\)-lexicase in comparison
to tournament selection.

Figure 2 also shows a visual representation of the described overall
performance benefit of using tournament selection instead of
\(\epsilon\)-lexicase selection for the chosen symbolic regression task.
It can be observed that at all stages of the evolution tournament
selection based GP achieves a lower mean error for both data sets.

@@

None of the differences in mean size are statistically significant.
Again, based on the results obtained in my experiment, the evidence
produced fails to reject the null hypothesis \(H0_{3}\) that selection
has no significant effect on differences in program size. It should be
noted that the configuration of both GP algorithm did include two
mechanisms that place limitations on the growth behavior of the
individual solutions to avoid excessive bloating (see sub section
3.2.4).

@@

The objective of this research project has been to answer the question
if the choice between two commonly used parent selection operators,
tournament selection and \(\epsilon\)-lexicase selection, significantly
influences the generalization behavior of GP for symbolic regression
tasks.

The results obtained from my experiment did not demonstrate any
statistically significant differences in generalization, both GP systems
achieve a slightly lower mean error if evaluated for the training data
sets than if evaluated for the unseen fitness cases of the testing data
set. The mean differences between testing and training errors did not
show statistical significance differences for both algorithms
(summarized in table 9).

To further explore differences in generalization behavior the mean for
both algorithms training and testing error over all generations for all
50 independent runs was visualized and analyzed (see figure 2). A very
small and probably insignificant gap did show for \(\epsilon\)-lexicase
selection in the final phase of the evolutionary search, in the grand
scheme of things this approach also did not yield any clear evidence of
differences in generalization.

A final attempt to explore possible differences between both GP systems
has been to explore differences in growth behavior as an indicator for
generalization. This idea was based on the wide spread hypothesis that
increases in overall program size might be an indication of over fitting
to the training data, the concept has been discussed in sub section 2.2.
Again, based on my results I could not find statistically significant
differences in the growth behavior between both selection operators.

An interesting finding of my analysis is, that the overall performance
of tournament selection has shown to be significantly better than that
of \(\epsilon\)-lexicase selection for the selected symbolic regression
problem. This finding was rather unexpected based on the research of La
Cava, Spector and Danai (2016) and La Cava \emph{et al.} (2017) that
primarily inspired this research project in the first place.

@@

\hypertarget{I}{%
\section*{I - References}\label{I}}
\addcontentsline{toc}{section}{I - References}

@@

\hypertarget{II}{%
\section*{II - Statutory Declaration}\label{II}}
\addcontentsline{toc}{section}{II - Statutory Declaration}

\includegraphics{./private/erklaerung.pdf}\\
